# PROSPER LOAN DATA ANALYSIS 
## By Emily Hazelton

------

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(GGally)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
loans_full_set <- read.csv('prosperLoanData.csv')

# Select variables of interest(instructions are to look at 10-15 variables)
loans <- select(loans_full_set, 
                LoanOriginalAmount, Term, LoanStatus, EmploymentStatus, 
                EmploymentStatusDuration,  CreditScoreRangeLower, 
                CreditScoreRangeUpper, DebtToIncomeRatio, StatedMonthlyIncome,  
                LP_CustomerPrincipalPayments, LP_InterestandFees, 
                EstimatedReturn)

loans <- rename(loans,
       loan_amount = LoanOriginalAmount,
       loan_term_months = Term,
       loan_status_current = LoanStatus,
       employment_status = EmploymentStatus,
       employment_duration_months = EmploymentStatusDuration,
       credit_score_lower = CreditScoreRangeLower,
       credit_score_upper = CreditScoreRangeUpper,
       debt_income_ratio = DebtToIncomeRatio,
       stated_income_monthly = StatedMonthlyIncome,
       principal_paid = LP_CustomerPrincipalPayments,
       interest_and_fees_paid = LP_InterestandFees,
       estimated_return = EstimatedReturn
       )

```

This report explores loan data provided by the online lender Prosper (a tidy 
dataset that Udacity prepared for student projects). The data spans from 
November, 2005 to March, 2014. Any column headers listed as "current" was 
current as of March 11, 2014, when the dataset was last pulled. 

# Univariate Plots Section

There are 113,937 observations in the full dataset with 81 variables.
I have picked 12 variables for this exploration, as detailed below. I've also 
shown a sample of the first 6 rows for each variable. 

```{r echo=FALSE, Univariate_Plots}
dim(loans)
names(loans)
```


```{r echo=FALSE}
head(loans)
```

Questions:

* What are the most popular categories for the loan status? 

```{r echo=FALSE}
ggplot(aes(x = loan_status_current), data = loans) + 
  geom_bar() +
  coord_flip()
```

It looks like the most commonly occurring loan statuses are current, completed, 
and charged off. 

Questions:

* What proportion of loans are charged off?
* What factors predict that the borrower will go into loan delinquency? 
* What is the distribution of loan amounts? 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=loan_amount), data = loans) +
  geom_histogram()
```

The default bin sizes here for loan amount makes it hard to see anything
meaningful. 

```{r echo=FALSE}
ggplot(loans, aes(loan_amount)) +
  geom_histogram(binwidth = 250) +
  coord_cartesian(xlim = c(0,25000))
```

Unsurprisingly, loan amounts cluster around the 500 or 1,000 marks.

Most frequent loan amounts: 

* $4,000
* $15,000
* $10,000
* $5,000

Question: 

* How long do loan terms last?

```{r echo=FALSE}
ggplot(aes(x=loan_term_months), data = loans) +
  geom_histogram(bins = 5) +
  scale_x_continuous(breaks = c(12,36,60))
```

The loan terms in this data set fall into three bins: 12, 36, or 60 months, with 
36 months being the most common. Since there are only 3 discrete values, I am 
going to change the vector type from integer to a factor, so I can graph it
categorically in the bivariate section. I would also find the length-of-loan 
variable easier to interpret as 1, 3, or 5 years.

```{r echo=FALSE}
#factor loan_term_months based on observations in previous graph
loans$loan_term_years <- factor(loans$loan_term_months / 12)
ggplot(loans, aes(loan_term_years)) +
  geom_bar()
```

Questions:

* What is the average credit score for borrowers? I want to create an average 
from credit_score_lower and credit_score_uppers so that I have one variable to 
use for analysis. 

```{r echo=FALSE}
#average the upper and lower credit scores so it's easier to visualize and 
#compare to other variables laater
loans <- mutate(loans, 
    credit_score_average = (credit_score_upper + credit_score_lower)/2)

ggplot(aes(x=credit_score_average), data = loans) +
  geom_histogram(binwidth = 10) +
  scale_x_continuous(limits = c(400, 900))
```

It looks like borrowers' credit scores are normally distributed, 
with the peak in the 685-695 bin. 

Questions: 

* Why are there gaps in this continusous data set? Are credit score algorithms
defined in such a way that every-other band of 10 digits is not a score where
consumers can rank? NOTE: I made graphs of both credit_score_lower and
credit_score_upper, and the same banding occurs. I removed them from this report
for the sake of brevity, but the effect is not coming from the fact that I 
averaged these two variables. 
* How much do borrowers make? NOTE: this variable is self reported, not verified
by Prosper.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=stated_income_monthly), data = loans) +
  geom_histogram()
```

There are definitely some extreme outliers in here that are making the bins
very unhelpful. 

```{r echo=FALSE}
ggplot(aes(x=stated_income_monthly), data = loans) +
  geom_histogram(binwidth = 500) +
  xlim(0,15000)
```

Monthly income appears to be normally distributed around a peak of about $3,000, 
with a long tail out into higher values. Since I'm used to interpreting income
on an annual scale, I am going to create a new value here named 
stated_annual_income. 

```{r echo=FALSE}
loans <- mutate(loans,
                stated_annual_income = stated_income_monthly * 12)

ggplot(loans, aes(stated_annual_income)) +
  geom_histogram(binwidth = 5000) +
  xlim(0,300000)
  
```

The normal distribution of income centers around the $30,000 to $50,000 range.

Question: 

* What are borrowers' employment statuses?

```{r echo=FALSE}
ggplot(loans, aes(employment_status)) +
  geom_bar() +
  coord_flip()
```

Question: 

* How long have borrowers been employed?

```{r echo=FALSE}
ggplot(loans, aes(employment_duration_months)) +
  geom_histogram(binwidth = 6, na.rm = TRUE) +
  xlim(0,400)
```

The above bins are 6-month time frames. I find this data difficult to interpret.
I am going to convert it to yearly numbers.

```{r echo=FALSE}
loans <- mutate(loans,
                employment_duration_years = employment_duration_months/12)

ggplot(loans, aes(employment_duration_years)) +
  geom_histogram(binwidth = .5) +
  xlim(0,40)

```

The most common bin is the .25-.75 years of employment, with the count of years 
employed steadily falling. Each bin above represents .5 years or 6 months.

Questions: 

* How much does Prosper make from its lending?

```{r echo=FALSE}
loans <- mutate(loans,
       bank_profit_current = principal_paid + interest_and_fees_paid - 
       loan_amount)
    
ggplot(aes(x = bank_profit_current), data = loans) +
  geom_histogram(binwidth = 500) + 
  xlim(c(-25000,10000))
``` 

I created a new variable called bank_profit_current. The formula is principal 
paid plus interest and fees paid minus the original loan amount. 

Questions:

*What are the unique loan_status values in the dataset? 
* How would this graph look for loans that are already closed?

```{r echo=FALSE}
#what are distinct factors in loans$loan_status_current?
unique(loans$loan_status_current)

closed_loans <- filter(loans, loan_status_current == 'Completed' | 
                         loan_status_current == 'Chargedoff')

ggplot(aes(x = bank_profit_current), data = closed_loans) +
  geom_histogram(binwidth = 500) + 
  xlim(c(-25000,10000))

```

The above graph shows bank_profit_current for only Completed or Chargedoff 
loans. It looks like closed loans are more scewed to positive profit for the 
bank than all loans (open and closed). This makes sense.

Questions:

* What is the average profit per closed loan?

Mean bank profit on closed loans: 

```{r echo=FALSE}
mean(closed_loans$bank_profit_current)
```

Average profit of $21.12 seems low to me.

Questions: 

* How does actual profit compare to the estimated profit caculated by Prosper
when they funded the loan? 
* How much data do I have in the closed_loans dataset? Could I use this dataset
for the rest of my analysis and have enough observations to draw conclusions 
that could help predict future performance of loans? 

```{r echo=FALSE}
dim(closed_loans)
```

Yes, over 50,000 observaions seems sufficient to reasonably trust any 
conclusions drawn from this dataset. From here forward, I will be looking at
this more specific dataset, named closed_loans in the R environment.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_status_current)) + geom_bar()
```

The above graph visually shows the counts of Chargedoff vs Completed loans. 
Completed loans are loans where the borrower completely paid off the original 
loan amount. 24.0 percent of closed loans are in the Chargedoff category. 

Questions: 
* How many 1, 3, and 5 year loans fall within the closed status?
* What is the spread of loan amounts within closed_loans?

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_term_years)) +
  geom_bar()
```

Three-year loans are still the  most common. In relation to the 3-year loans, 
there are proportionally more 1-year loans and less 5-year loans as compared
to the full Prosper loan dataset. 

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount)) +
  geom_histogram(binwidth = 250) +
  coord_cartesian(xlim = c(0,25000))
```

The spread for closed loans leans more towards smaller loans than the full
Prosper set. Perhaps consumers were asking for smaller loans historically, or
perhaps Prosper was more conservative in how much funding they would award to 
consumers as they refined their profit predictions -- I don't have a way to find
the answer to that question with the dataset that I have. 

Most popular loans:

* $4,000
* $5,000
* $3,000
* $2,000

With these smaller loan sizes, this smaller profit makes more sense. This data
was also collected over the span of the recession, so interest rates would have
been very low during that time. That also helps explain the low ROI per loan. 

Question: 

* What type of return does the bank expect to make on its lending?

```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(estimated_return)) %>%
  ggplot(aes(estimated_return)) +
    geom_histogram(binwidth = .01) +
    coord_cartesian(xlim = c(-0.1,0.25))
```

This graph shows what the bank estimated the return to be when they funded the
loan. It looks like there is a spread between 0.025 and 0.2 (or 2.5 to 20 
percent of return on investment).

Questions: 

* What does actual return look like? 

```{r echo=FALSE}
closed_loans <- mutate(closed_loans,
    actual_return = bank_profit_current / loan_amount)

ggplot(closed_loans, aes(actual_return)) + 
  geom_histogram(binwidth = .025) 
```

Most loans are profitable for the bank, with more common profits ranging from 
just breaking even (.0) to 50 percent (.5). There is a strange peak above 0.5,
and I'm not sure why that is. If I were working for Prosper, I would definitely
look at that. But for the sake of this project, I'm going to move past it. 

The formula I used for acual return was bank_profit_current divided by 
loan_amount. 

Question:

* Is Prosper's model for predicting profit conservative? Or pretty accurate?

```{r echo=FALSE}

#difference <- closed_loans$actual_return - closed_loans$estimated_return

#head(difference, 10)
# graph isn't working -- need to figure out why there are na values in result
# I didn't think I had na values in actual_return
# ggplot(difference, aes(difference)) + geom_histogram()
```

* What is the distribution of debt-to-income ratio for borrowers?

```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(debt_income_ratio)) %>%
  ggplot(aes(debt_income_ratio)) +
    geom_histogram(binwidth = 0.5)
```

And zoomed in to remove that outlier column:

```{r echo=FALSE}
closed_loans %>%
  filter(!is.na(debt_income_ratio)) %>%
  ggplot(aes(debt_income_ratio)) +
    geom_histogram(binwidth = 0.05) +
    xlim(0,1.5)
```

The graph peaks at the 0.125-0.175 bin (or 12.5 to 17.5 percent debt-to-income
ratio).

# Univariate Analysis

### What is the structure of your dataset?

I decided to focus my investigation on loans that have been completed or 
charged off. There are 50,066 observations that meet this criteria.

* In the closed_loans dataset, there are 38,074 completed loans and 11,992
charged off loans. 24.0 percent of closed Prosper loans were charged off.
* Prosper loans are one, three, or five years in duration. 
* The most common credit score for borrowers is between 665 and 725.
* The average profit per loan is $21.12.
* The loan amount for at least half of Prosper loans in the closed_loan dataset 
is equal to or less than $5,000.

### What is/are the main feature(s) of interest in your dataset?

For the rest of this analysis, I would like to focus on what variables can 
help predict actual_return. 
 
I anticipate that the following variables will have the strongest correlation
to actual_return: 

* credit_score_average
* loan_term_years
* loan_amount
 
### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I will also explore what covariation exists between actual_return and the 
following variables. I expect less of a correlation amongst these, but want to 
check to see if any of these have a higher impact than I would expect: 

* employment_status
* debt_income_ratio
* employment_duration_years
* stated_annual_income


### Did you create any new variables from existing variables in the dataset?

Yes. I transformed several variables into forms that I feel will be more concise
for visualization purposes:

* credit_score_average: calculated from credit_score_lower and 
credit_score_upper
* loan_term_years: loan_term_months divided by 12; also converted to a factor
* employment_duration_years: employment_duration_months divided by 12 
* stated_annual_income: stated_income_monthly multiplied by 12 

I wanted one variable to use for credit score, rather than two. And I find it 
easier to interpret the data when the loan term, employment duration, and income
is stated annually rather than monthly. 

Now, I have the following variables to move into the bivariate analysis: 

```{r echo=FALSE}
# Re-order columns to make more sense logically; drop variables that I am not
# going to look at further.

closed_loans <- select(closed_loans, 
                       loan_status_current, loan_term_years, estimated_return, 
                       loan_amount, principal_paid, interest_and_fees_paid, 
                       bank_profit_current, actual_return, employment_status,
                       employment_duration_years, stated_annual_income, 
                       credit_score_average, debt_income_ratio)

dim(closed_loans)
```

```{r echo=FALSE}
names(closed_loans)
```

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

See above section for changes made to the dataset.

# Bivariate Plots Section

Let's start by looking at the correlations between different variables. 
```{r echo=FALSE}
ggcorr(closed_loans, label = TRUE)
```

It looks like there may be a negative correlation between annual income and 
estimated return. My primary factor of interest is estimated_return (displayed
across the bottom). Some of the strong correlations -- between principal paid
and loan_amount or actual_return and bank_profit_current -- are obvious because
the variables do have an explicit relationship. 

Question:

* How is actual_return related to credit_score_average?

```{r echo=FALSE, Bivariate_Plots}
ggplot(closed_loans, aes(credit_score_average,actual_return)) +
  geom_point(alpha = .05) +
  xlim(400,900) +
  geom_hline(yintercept=0, color = 'red', size = 1)
```

It doesn't look like I will be able to use credit score in a predictive model
for actual return. I really expected more of a correlation. It does look like
individuals who have a score over 800 are a very safe bet for investment.
Perhaps the overplotting on this graph is hiding a trend that is there. 

Question: 

* Would a grouped boxplot show a better trend? 

```{r echo=FALSE}
ggplot(closed_loans, aes(credit_score_average, actual_return)) +
  geom_boxplot(aes(group = cut_width(credit_score_average,10))) +
  coord_cartesian(xlim = c(440,880), ylim = c(-.75,1)) +
  geom_hline(yintercept = 0, color='red', size=1) 
```

It looks like the median return is above 0 for all ranges except those
centering at 450 and below. And the peak in the median return occurs at a 
credit score of 570. 

With this graph, it looks like there is a pattern in the relationship between
credit score and return. Those with poor or fair credit scores are bringing the
highest rate of return. This is likely because these individuals pay higher 
interest rates. 

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_term_years,actual_return)) +
  geom_boxplot()
```

It looks like three-year loans have the highest median percentage of return. And
it looks like Prosper is losing the most money on five-year loans. In fact, 
almost half of five-year loans are losing money.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount, actual_return)) +
  geom_point(alpha = .05) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

Here we see the effects of overplotting within the bands of most common loan
amounts. It does look like the return falls much more above 0, which is good
for Prosper. Let's take a look at this with a boxplot grouping to see where 
the medians lie by $1,000 bins.

```{r echo=FALSE}
ggplot(closed_loans, aes(loan_amount,actual_return)) +
  geom_boxplot(aes(group = cut_width(loan_amount, 1000))) +
  xlim(0,25500) +
  coord_cartesian(ylim = c(-.5,.5)) +
geom_hline(yintercept=0, color = 'red', size = .75)
```

Return is above 0 for all loan amounts. There really is no evidence of variation
in the loan amount changing the ROI. 

Question: 

* How does debt-to-income-ratio impact return?

```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

The relationship between bank profit and stated monthly income doesn't seem
to have a linear pattern, but an outlier is seriously skewing the x scale of 
that graph, so I'm going to zoom in.

```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return)) +
  geom_point(alpha = .1) +
  coord_cartesian(xlim=c(-.25,2.5)) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

It seems as though a higher debt_income_ratio (DIR) is related to a higher 
profit. There seem to be some tiers to the data, althought it is still hard to see 
due to overplotting. But it seems that there are jumps in the level of profit at
both 0.25 and 0.5 marks for DIR.

Question: 

* How does employment status impact return?

```{r echo=FALSE}
ggplot(closed_loans, aes(reorder(employment_status,actual_return, FUN = median), 
  actual_return)) +
  geom_boxplot() +
  coord_flip() 
```

These are some strange tiers, with redundancy. My guess is that the option set 
changed at some point in data collection. For example "employed" was later 
changed to two options -- "part-time" and "full-time."

```{r echo=FALSE}
closed_loans %>%
  #ignore fields that are: blank, 'Not available,' and 'Other'
  filter(employment_status == 'Employed' | 
         employment_status == 'Full-time' |
         employment_status == 'Not employed' |
         employment_status == 'Part-time' |
         employment_status == 'Retired' |
         employment_status == 'Self-employed'   ) %>%
  ggplot(aes(actual_return, y = ..density..)) +
    geom_freqpoly(aes(color=employment_status),binwidth=.025)
```

The plot shows density, so that the count of each category doesn't scew the 
graph. The graph shows a proportion for each category.

It is surprising to me that those with an employment_status of 'employed' bring
Prosper a lower actual_return. However, I know that some of the data collection
criteria changed in 2009, based on the data dictionary. I don't know exactly
when the employment_status optionset would have changed. But if 'Employed' is 
from an earlier date, it's not a good comparison (because there are many 
environmental differences that could make a difference if the data was collected
during two different time periods).

Let's zoom in to the peak area of this graph.

```{r echo=FALSE}
closed_loans %>%
  #ignore fields that are: blank, 'Not available,' and 'Other'
  filter(employment_status == 'Employed' | 
         employment_status == 'Full-time' |
         employment_status == 'Not employed' |
         employment_status == 'Part-time' |
         employment_status == 'Retired' |
         employment_status == 'Self-employed'   ) %>%
  ggplot(aes(actual_return, y = ..density..)) +
    geom_freqpoly(aes(color=employment_status),binwidth=.025) +
    xlim(c(0,0.5))
```

It's interesting that retired and part-time individuals seem to peak at a higher
actual_return rate. Perhaps it is a similar phenomenon to why Prosper makes a 
better return for those with lower credit scores -- perhaps these individuals 
are assigned a higher interest rate. 

Question:

* How does employment duration impact return?

```{r echo=FALSE}
ggplot(closed_loans, aes(employment_duration_years,actual_return)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

```{r echo=FALSE}
ggplot(closed_loans, aes(employment_duration_years,actual_return)) +
  geom_boxplot(aes(group = cut_width(employment_duration_years,1))) +
  xlim(0,20) +
  ylim(-.5,.5) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

Question:

* How does stated annual income impact return?

```{r echo=FALSE}
ggplot(closed_loans, aes(stated_annual_income,actual_return)) +
  geom_point(alpha = .1) +
  xlim(0,200000) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```
 
```{r echo=FALSE}
ggplot(closed_loans, aes(stated_annual_income,actual_return)) +
  geom_boxplot(aes(group = cut_width(stated_annual_income,10000))) +
  xlim(0,200000) +
  ylim(-.5,.5) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

## Relationships between other variables

Question: 

* What is the relationship between credit score and debt_income_ratio?

```{r echo=FALSE}
ggplot(closed_loans, aes(factor(credit_score_average), debt_income_ratio)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0,1)) +
  theme(axis.text.x = element_text(angle = 90))

```

It looks like debt-to-income ratio is higher in the middle regions of the 
credit score range. I believe that debt to income ratio is one factor that
influences credit score, so it makes sense that those with an excellent credit
score would have a low debt to income ratio. I wonder if those with a score 
under 530 simply have a harder time getting a loan and therefore don't have as 
much debt.

Question: 

* Is a longer employment history correlated with a better credit score?

```{r echo=FALSE}
closed_loans %>%
  filter(employment_status == 'Employed' | employment_status == 'Full-time' |
           employment_status == 'Part-time' | 
           employment_status == 'Self-employed') %>%
  ggplot(aes(credit_score_average, employment_duration_years)) +
    geom_point(alpha = .1) +
    xlim(450,850)
```

A longer employment duration does not appear to correlate with a better 
credit score.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Loan term covaried with actual return. The return for one-year loans was mostly
positive, but not much more than 0. Three-year loans brought in the highest 
return. Five-year loans were responsible for the most lost in return. 

Credit score also varied with actual return. Credit scores centering right under 
600 actually brought the highest level of return. 

Otherwise, I really didn't find any other patterns of covariation. 

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I didn't find any areas of covariation in the non-central variables that I did 
explore. 

### What was the strongest relationship you found?

Credit score and actual return. Which makes sense, since the industry has spent
years perfecting a single indicator that will tell the bank how trustworthy you
are as an investment. 

# Multivariate Plots Section

Question:

* Does loan term in years add any meaningful pattern to a graph of actual_return
and credit_score_average?

```{r echo=FALSE, Multivariate_Plots}
ggplot(closed_loans, aes(factor(credit_score_average), actual_return, fill=factor(loan_term_years))) +
  geom_boxplot() +
  scale_fill_brewer(type='seq', guide=guide_legend(title='Loan Term')) +
  theme(axis.text.x = element_text(angle = 90))
```

Within loans that are three years in duration (the most commonly occurring and
easiest to see), it looks like the median return is highest in the upper 500s 
and lower 600s. The median return steadily increases above a credit score of 
630.

Question: 

* Does employment_status help add a pattern to the relationship between credit
score and return? 

```{r, echo=FALSE}
ggplot(closed_loans, aes(credit_score_average,actual_return,
                         color=employment_status)) +
  geom_point(alpha = .5) +
  xlim(400,900) +
  geom_hline(yintercept=0, color = 'red', size = .75)
```

In this graph, we can see that the colors associated with 'not available,' 'not
employed,' and 'other' occur more frequently in the lower credit scores, and 
that employment of some type or retirement occurs more frequently in the higher 
credit score ranges. 

Question:

* Does debt-to-income ratio combined with employment status offer any new 
insight?

```{r echo=FALSE}
ggplot(closed_loans,aes(debt_income_ratio,actual_return,
                        color=employment_status)) +
  geom_point(alpha = .5) +
  coord_cartesian(xlim=c(-.25,2.5)) + stat_ellipse()
```

Adding employment status to this graph doesn't show any new insights. Debt-to-
income ratio combined with empoyment status does not seem to build a pattern.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

It looks like those with employment tend to have higher credit scores. But that
was not particularly relevant to my exploration with actual return. I really 
struggled to uncover any interesting new insights with the multivariate 
analysis. 

### Were there any interesting or surprising interactions between features?

Not really.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(closed_loans, aes(actual_return)) + 
  geom_histogram(binwidth = .025) +
  labs(title = 'Return on Investment for Loan Funding', x='Actual Return*', 
       caption='*Actual return is profit as a percentage of initial loan 
       amount.', y='Count')
```

### Description One

This plot shows that most loans that have a positive return on investment. The 
average return for a funded loan is 2.8 percent. 

```{r echo=FALSE}
summarize(closed_loans,
          mean_return_percent = mean(actual_return)*100,
          median_return_percent = median(actual_return)*100)
```

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(closed_loans, aes(loan_term_years, actual_return)) +
  geom_boxplot() + 
  labs(title ='Actual Return by Loan Duration', x='Loan Term (Years)', 
       y='Actual Return*', caption = '*Actual return is profit as a percentage 
       of initial loan amount.')
```

### Description Two

Prosper gives out loans that are one, three, or five years in duration. Almost 
90 percent of loans are provided with a three year term. The median return 
for these loans is 13.6 percent. However, the average return is only 4 percent.
Negative outliers are bringing the average lower than the median. 

The median return for one-year loans is 5.9 percent, with average profit at 3.8
percent. And the median return for five-year loans is 5.8 percent, with average
profit at negative 13.4 percent. As can be seen in the graph above, the second
quartile for five-year loans falls into a negative return. The five-year loan
category is not profitable for Prosper. 

```{r echo=FALSE}
closed_loans %>%
  group_by(loan_term_years) %>%
  summarize(
    count = n(),
    return_mean_percent = mean(actual_return, na.rm = TRUE)*100,
    return_median_percent = median(actual_return, na.rm = TRUE)*100
  )
```

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(closed_loans, aes(credit_score_average, actual_return)) +
  geom_boxplot(aes(group = cut_width(credit_score_average,20))) +
  coord_cartesian(xlim = c(440,880), ylim = c(-1,1.25)) +
  geom_hline(yintercept=0, color = 'red', size = .75) + 
  labs(title ='Actual Return by Borrower Credit Score', x='Credit Score', 
    y='Actual Return*', caption = '*Actual return is profit as a percentage of 
    initial loan amount.')
```

### Description Three

This graph shows that the median profit is above 0 for almost all credit score
categories except those below 470. However, if we look at the mean profit, 
return becomes reliable above a credit score of 530.

```{r echo=FALSE}
closed_loans %>%
  group_by(cut_width(credit_score_average,20)) %>%
  summarize(
    count = n(),
    return_mean_percent = mean(actual_return, na.rm = TRUE)*100,
    return_median_percent = median(actual_return, na.rm = TRUE)*100
  ) %>% rename("bins" = "cut_width(credit_score_average, 20)")

```

------

# Reflection

What was challenging: 

* I was surprised that I didn't uncover a clear linear model or strong pattern 
between actual_return and the various factors I investigated. I was frustrated 
about that. Staying motivated on this project was difficult, because I didn't
feel like I uncovered any key insights. 
* I didn't get a chance to build a model because there were no linear 
relationships that emerged. Perhaps a more complex model could work, but I am 
currently not educated on those, so it was beyond the scope of my experience at 
this point in time.

What went well:

* I learned a lot about ggplot, using R studio, and how to approach the EDA
process. One example of a process technique that I learned is to do a thorough 
exploration about variation (i.e. univariate analysis) and then move on to 
explore covariation (bivariate and mutlivariate analysis).
* I enjoyed the creativity and curiosity that the EDA process requires. I was
able to practice using a wide range of graph types in ggplot.
* I am excited to take this skillset to analyzing other datasets.

Further areas for investigation: 

* How does early payoff impact actual return? That is, if a borrower pays their
loan off early, I anticipate that actual return would be lower.  
* Analyze the difference between estimated return and actual return. Focus on 
areas where the bank anticipated profit but lost money. Is there a factor that
their predictions should also take into account? Or is that just part of the 
risk of the lending process (e.g. the exploration would not uncover additional 
factors to create better predictive models of expected profit)? 

#References

I consulted the following resources to help with this EDA project: 

* R for Data Science by Hadley Wickham & Garrett Grolemund
* Online support forums for specific error messages